---
title: "Chapter6"
author: "Pekka Tolli"
date: "7 December 2018"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)

```


#Week 6 - Longitudinal set of rats featuring pipe(r) of Hameln

This is the last IODS exercise for now. Time for longitudinal analysis of rats and humans, featuring some pipe(r) functions (stupid pun intended). 

## Part 1 RATS:
On this first part  I'll analyse the longitudinal rats data set by Kimmo Vehkalahti from a nutrition study on three groups of rats. The data each rat's weight being recorded once a week for 9 weeks. Thus the data is longitudinal, i.e. it tracks each rat over a time period. I will test if the nutrition changes have impact on the weight change of a rat. 


````{r}
RATSL <- read.csv("C:/Users/pekka/Documents/GitHub/IODS-project/Data/rats.txt", sep  ="\t", header = T)

RATSL$Group <- factor(RATSL$Group)
RATSL$ID <- factor(RATSL$ID)

str(RATSL)
dim(RATSL)
summary(RATSL)
head(RATSL)
```

We have data set with 5 variables and 176 rows containing observations of 16 rats across 9 weeks of time. 

##Visualizing RATS data
Let's draw basic line charts of the unstandardized RATS data classified by nutrition group: 

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

ggplot(RATSL, aes(y=Weight, x=factor(Time), fill=Group)) +
  geom_boxplot()

```
A couple of observations emerge: First, practically all rats have increased their weight during the study. Second, group 1 rats were lighter at the beginning and the end of the study. Third, group 2 has a potential outlier (a big rat). 


##Standardizing the RATS data
To do proper analysis, let's standardize the data:
```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(standard_RATSL = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)

```

And let's see again how the standardized data looks like:

````{r}
ggplot(RATSL, aes(x = Time, y = standard_RATSL, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rats")

ggplot(RATSL, aes(y=standard_RATSL, x=factor(Time), fill=Group)) +
  geom_boxplot()

```

We now can see that Y-axis with weights has now standardized scale. Variability inside the groups was not removed but we made data more suitable for analysis


```{r}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
```

````{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)") +
  ggtitle("RATS: means and standard errors")
````

The plot above represents the mean and standard deviations of weight of each three groups over time. As discussed, group 1 has the lowest mean of weight, while Groups 2 and 3 are closer to each other.

##Removing outliers

Lets see if the data has any outliers. If yes, remove them. 
````{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL8S)

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "green") +
  scale_y_continuous(name = "mean(Weight)")
````
In group 2 there is one observation quite far off from the others as pointed out in the initial observation. Let's remove it

```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "green") +
  scale_y_continuous(name = "mean(Weight, no outliers)")

````
Now the data looks better, no clear outliers anymore.

Time for checking the real differences between the groups. It looks from the plots that groups behaved bit differently but let's see it with ANOVA.
First, add the baseline from the original data as a new variable to the summary data



````{r}
RATSL8S1 <- RATSL8S %>%
  mutate(baseline = RATSL$WD1)

str(RATSL)
str(RATSL8S)
str(RATSL8S1)
str(RATSS)

Create a  linear model with the mean as the response 
#fit <- lm(mean ~ baseline + Group, data = RATSL8S1)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

````


RATSLS2 <- RATSS_B %>%
  mutate(baseline = filter(RATSL, Time==1)$Weight)
RATSLS2$mean <- as.numeric(RATSLS2$mean)

str(RATSLS2)




#Part 2: BPRS

#Let's then analyse the BPRS data, which is about the 
#The long format of BPRS (BPRSL) has 360 rows and 5 columns i.e. variables.

``` {r}
BPRSL <- read.csv("C:/Users/pekka/Documents/GitHub/IODS-project/Data/bprs.txt", sep  ="\t", header = T)

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

str(BPRSL)
dim(BPRSL)
summary(BPRSL)
head(BPRSL)
```


Let's visualize the data:
``` {r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

###Linear regression model

Let's proceed to the linear model assuming that measurements are independent. That would be a basic regression model run by lm() function in R. Here we probably do not need to remove the outliers as in rats dataset, as lm is not so sensitive to them as ANOVA is.

